Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-11 sftrends_manu.Rnw
text for observed trends added

source('R/funcs.r')
library(WRTDStidal)
library(tidyverse)
library(gridExtra)
library(grid)
library(Hmisc)
library(ggrepel)
library(ggtree)
library(NADA)
@

@@

\begin{document}
% \linenumbers

@@

The goal of this study was to provide a comprehensive description of nutrient trends in the Delta that will inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay. We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production. Although quantitative descriptions of change can be ends in themselves, the results were expected to have greater impact as a means to more detailed understanding of ecosystem properties.  Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects. Overall, this analysis is expected to further an ecological and management-based understanding of dynamics in San Francisco Bay, with implications for water quality restoration and protection of this prominent system.

@@

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.  Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

A total of twenty-seven \ac{wrtds} models were created, one for each nitrogen analyte at each station.  The functional form of \ac{wrtds} is a simple regression\cite{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.  However, observed seasonality patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends from 1976-1995 and 1996-2013 varied considerably between sites and analytes.  Overall,

kendall tests - observed, \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}

What about flow? \cref{fig:dinc10}

@@

\cref{fig:clmchl}, \cref{fig:d7c10trnds}, \cref{tab:d7chg}, \cref{tab:c10chg}

>>>>

source('R/funcs.r')
library(WRTDStidal)
library(tidyverse)
library(gridExtra)
library(grid)
library(Hmisc)
library(ggrepel)
library(ggtree)
library(NADA)
library(english)
@

@@

\begin{document}
\linenumbers

@@

The goal of this study was to provide a comprehensive description of nutrient trends in the Delta to inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay. We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production. Although quantitative descriptions of change can be ends in themselves, the results were expected to have greater impact as a means to more detailed understanding of ecosystem properties.  Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects. Overall, this analysis is expected to further an ecological and management-based understanding of dynamics in San Francisco Bay, with implications for water quality restoration and protection of this prominent system.

@@

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.  Quantitative descriptions of nutrient dynamics in the Delta are challenging given many nutrients sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

A total of thirty \ac{wrtds} models were created, one for each nitrogen analyte at each station.  The functional form of \ac{wrtds} is a simple regression\cite{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick29}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.  However, observed seasonality patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).

@@

<<echo = F, cache = T>>=
data(trnds_seasyr)

trndsyr <- filter(trnds_obs, cat %in% c('1976-1995', '1996-2013')) %>% 
  select(Location, Site_Code, resvar, cat, ztest, perchg)

obsdir <- filter(trndsyr, ztest < 0.05) %>% 
  mutate(dirchg = ifelse(perchg < 0, 'down', 'up')) %>% 
  group_by(cat, resvar) %>% 
  summarise(
    up = sum(dirchg == 'up'), 
    down = sum(dirchg =='down'),
    tots = sum(up, down),
    up = as.character(english(up)),
    down = as.character(english(down)),
    tots = as.character(english(tots))
  ) %>% 
  as.data.frame %>%
  split(., .$resvar)
P8dec <- filter(trndsyr, 
    Site_Code %in% 'P8' &
    resvar %in% 'nh'
  ) 
@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).  More sites had decreasing trends for the observed data from 1996-2013.  \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing). Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.  For example, P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites. Trends by season were similar such that increases were generally observed in all seasons for the first time period from 1976-1995 and decreases were observed for the second (\cref{fig:trndcomp2,fig:trndcomp3}), although trends were noisier and significant changes were less common compared to the annual comparisons.

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are potentially confounded by flow effects.  A simple monotonic change over time could reflect moblization or dilation effect of flow rather than an empirical system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data. YThe raw data are presented in \cref{fig:dinc10}a and the annual predicted and flow-normalized results form \ac{wrtds} are shown in \cref{fig:dinc10}b.

@@

\cref{fig:clmchl}, \cref{tab:d7chg}

### 2017-01-11 sftrends_manu.tex
text for observed trends added

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% \linenumbers

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.62$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.72$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.  However, observed seasonality patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends from 1976-1995 and 1996-2013 varied considerably between sites and analytes.  Overall,

kendall tests - observed, \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}

What about flow? \cref{fig:dinc10}

>>>>

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick29}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.62$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.72$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.  However, observed seasonality patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).  More sites had decreasing trends for the observed data from 1996-2013.  Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing). Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.  For example, P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites. Trends by season were similar such that increases were generally observed in all seasons for the first time period from 1976-1995 and decreases were observed for the second (\cref{fig:trndcomp2,fig:trndcomp3}), although trends were noisier and significant changes were less common compared to the annual comparisons.

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are potentially confounded by flow effects.  A simple monotonic change over time could reflect moblization or dilation effect of flow rather than a response to nutrient soruces.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data. YThe raw data are presented in \cref{fig:dinc10}a and the annual predicted and flow-normalized results form \ac{wrtds} are shown in \cref{fig:dinc10}b.
